**NOTE: I mistakenly deleted all of notes and writing from the previous project assignment. I did completely change and reformat most of my data, but I will make sure not to delete anything from here on out.**

4/22: Since my hypothesis is that a higher fastball velocity does NOT increase the likelihood of a pitcher needing Tommy John surgery, I know that I will need to choose statistical analysis tests that take multiple means into account at once. I know that a two-sample t-test is a common way to do this since it can compare two separate samples and whether or not they differ from each other. In my case, a two-sample t-test will test to see if there is a significant difference between pitchers who did and pitchers who did not have surgery. It wouldn't work for me to use an ANOVA test because that test is typically used when there are three or more sample groups and in my case I only have two (TJ vs no TJ). I know the test-t will be a good test but not sure if there are any other statistical analysis's I could use.
```{r}
library(tidyverse)
```

```{r}
tj_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRza4COxnPzAC9B4iLHSftyQLGLauOseQDMyTZe1Rmv14tBvApecwU8JnCBJ-uGqiXqorOtIeThPoVK/pub?gid=0&single=true&output=csv")

no_tj_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRiEXf3DkWF1JMN1XLNW0UaPR3pKiNVmnQFfFP_RHdihRQU9WVRxp0Uf_a7oxWBrExJYmvDNvCqvE5w/pub?gid=0&single=true&output=csv")

velo_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTH6AIJrKc_Hzu4HbBbNwn54krNj_7b8Pmh-wTAM8AjGSoY6bphzb_KEjN_U7QKZj8HKoUwJ778yT9w/pub?gid=0&single=true&output=csv")

```

```{r}
tj_data <- tj_data |> 
  mutate(had_TJ = 1)

no_tj_data <- no_tj_data |> 
  mutate(had_TJ = 0)
```
I used AI to give me recommendations for how to best clean up and combine my data so that I didn't have to have three separate files with data. I used the mutate functions to add a 0 to all the pitchers who have not had TJ and a 1 to everyone who has.

```{r}
all_pitchers <- bind_rows(tj_data, no_tj_data)
```
I deleted all of the unnecessary info from the google sheets and then looked up the best way to combine data within R studio. Using the bind_rows function, I combined the lists of pitcher who have and have not had surgery into one data frame. Since I already used the mutate function, each pitcher now has a 0 or 1 next to their name depending on if they had surgery.

```{r}
velo_for_all <- all_pitchers |> 
  left_join(velo_data |>  select(Name, ff_avg_speed), by = "Name")
```
With my newly combined dataframe, I also added a column for every pitcher that pulled their pitch speed from my original google sheet that had the velo for every pitcher. For pitchers that have N/A listed for their velo, I will manually enter that data if they are selected when I use the set.seed function to get my random sample. I think that I forgot remove the qualifier for number of innings pitched when originally pulling my data from baseballsavant, meaning that some of the pitchers who have had surgery didn't pitch enough innings before getting hurt so their data wasn't listed on the velocity table.

```{r}
# Histogram for TJ pitchers
ggplot(merged_data |>  filter(had_TJ == 1), aes(x = ff_avg_speed)) +
  geom_histogram(binwidth = 1, fill = "#D55E00", color = "white", alpha = 0.85) +
  labs(
    title = "Fastball Velocity: Tommy John Pitchers",
    x = "Average Fastball Velocity (mph)",
    y = "Number of Pitchers"
  ) +
  theme_minimal()

# Histogram for Non-TJ pitchers
ggplot(merged_data |>  filter(had_TJ == 0), aes(x = ff_avg_speed)) +
  geom_histogram(binwidth = 1, fill = "#009E73", color = "white", alpha = 0.85) +
  labs(
    title = "Fastball Velocity: Non-Tommy John Pitchers",
    x = "Average Fastball Velocity (mph)",
    y = "Number of Pitchers"
  )

```
I created a histogram for each sample group that I have just to help me better visualize the data.
