Gibson Hirt

### 4/29
This week, I ran two types of statistical analysis tests -- a t-test and a logistical regression test like you recommended. I also followed your advice of running these test both for the entire populations and with random samples meaning that in total, I ran four separate statistical analyses. Starting with task 7 (line 71) I created random samples for both sets of data -- players who have and players who have not had Tommy John surgery. Line 84 is all of my notes from the random sample process and what I am doing to deal with pitchers who have N/A listed as their average pitch speed. I successfully ran two t-tests and was surprised to find drastically different results between the two. My test of the random samples resulted in a p-value of 0.9453, implying that the correlation between fastball velocity and surgery is weak and I should not reject my null hypothesis. Based on this test, I assumed that there is not a significant difference between the fastball velocity of pitchers who have had surgery vs pitchers who have not had surgery. However, when I ran the t-test with the full populations, I observed a p-value of 0.006698, allowing me to reject my null hypothesis and conclude that there is a significant difference in pitch speed from pitchers who have had surgery to those who have not.

I then ran both logistical regression tests for my data and observed similarly split results. The test using full populations resulted in a low p-value and positive ff_avg_speed coefficient. The positive coefficient indicates that as velocity increases, the likelihood of surgery does as well. The low p-value indicates that there is strong evidence to back up this claim. However, the test using the random sample did not come to the same conclusions. There was a negative ff_avg_speed coefficient, meaning that the data actually suggests that higher velocity actually decreases the likelihood of surgery. However, the p-value is extremely high at 0.943, meaning that there is not significant evidence that fastball velocity is associated with surgery within this sample.


### 4/22
Since my hypothesis is that a higher fastball velocity does NOT increase the likelihood of a pitcher needing Tommy John surgery, I know that I will need to choose statistical analysis tests that take multiple means into account at once. I know that a two-sample t-test is a common way to do this since it can compare two separate samples and whether or not they differ from each other. In my case, a two-sample t-test will test to see if there is a significant difference between pitchers who did and pitchers who did not have surgery. It wouldn't work for me to use an ANOVA test because that test is typically used when there are three or more sample groups and in my case I only have two (TJ vs no TJ). I know the test-t will be a good test but not sure if there are any other statistical analysis's I could use.

### Task 1: Load the tidyverse and print hypothesis
```{r}
library(tidyverse)

null_hypothesis <- "A higher velocity fastball does not make a MLB pitcher more likely to need Tommy John surgery."
alternate_hypothesis <- "A higher velocity fastball makes a MLB pitcher more likely to need Tommy John surgery."
```

### Task 2: Load data
```{r}
tj_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRza4COxnPzAC9B4iLHSftyQLGLauOseQDMyTZe1Rmv14tBvApecwU8JnCBJ-uGqiXqorOtIeThPoVK/pub?gid=0&single=true&output=csv")

no_tj_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRiEXf3DkWF1JMN1XLNW0UaPR3pKiNVmnQFfFP_RHdihRQU9WVRxp0Uf_a7oxWBrExJYmvDNvCqvE5w/pub?gid=0&single=true&output=csv")

velo_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTH6AIJrKc_Hzu4HbBbNwn54krNj_7b8Pmh-wTAM8AjGSoY6bphzb_KEjN_U7QKZj8HKoUwJ778yT9w/pub?gid=0&single=true&output=csv")

```
### Task 3: Add column to differentiate data
```{r}
tj_data <- tj_data |> 
  mutate(had_TJ = 1)

no_tj_data <- no_tj_data |> 
  mutate(had_TJ = 0)
```
I used AI to give me recommendations for how to best clean up and combine my data so that I didn't have to have three separate files with data. I used the mutate functions to add a 0 to all the pitchers who have not had TJ and a 1 to everyone who has.

### Task 4: Combine data
```{r}
all_pitchers <- bind_rows(tj_data, no_tj_data)
```
I deleted all of the unnecessary info from the google sheets and then looked up the best way to combine data within R studio. Using the bind_rows function, I combined the lists of pitcher who have and have not had surgery into one data frame. Since I already used the mutate function, each pitcher now has a 0 or 1 next to their name depending on if they had surgery.

### Task 5: Add pitch speed
```{r}
velo_for_all <- all_pitchers |> 
  left_join(velo_data |>  select(Name, ff_avg_speed), by = "Name")
```
With my newly combined dataframe, I also added a column for every pitcher that pulled their pitch speed from my original google sheet that had the velo for every pitcher. For pitchers that have N/A listed for their velo, I will manually enter that data if they are selected when I use the set.seed function to get my random sample. I think that I forgot remove the qualifier for number of innings pitched when originally pulling my data from baseballsavant, meaning that some of the pitchers who have had surgery didn't pitch enough innings before getting hurt so their data wasn't listed on the velocity table.

### Task 6: Create histograms
```{r}
# Histogram for TJ pitchers
ggplot(velo_for_all |>  filter(had_TJ == 1), aes(x = ff_avg_speed)) +
  geom_histogram(binwidth = 1, fill = "#D55E00", color = "white", alpha = 0.85) +
  labs(
    title = "Fastball Velocity: Tommy John Pitchers since 2020",
    x = "Average Fastball Velocity (mph)",
    y = "Number of Pitchers"
  ) +
  theme_minimal()

# Histogram for Non-TJ pitchers
ggplot(velo_for_all |>  filter(had_TJ == 0), aes(x = ff_avg_speed)) +
  geom_histogram(binwidth = 1, fill = "#009E73", color = "white", alpha = 0.85) +
  labs(
    title = "Fastball Velocity: Non-Tommy John Pitchers since 2020",
    x = "Average Fastball Velocity (mph)",
    y = "Number of Pitchers"
  )

```
I created a histogram for each sample group that I have just to help me better visualize the data. Non-surgery pitchers have a normal, unimodal distribution with almost no outliers. Surgery pitchers have a bimodal distibution with more spread than the other dataset.

### Task 7: Create random samples
```{r}
set.seed(8)

tj_pitchers <- velo_for_all |> filter(had_TJ == 1)
no_tj_pitchers <- velo_for_all |> filter(had_TJ == 0)

# I made my random sample sizes approximately 10% of the full population size
tj_sample <- tj_pitchers |> sample_n(15)
no_tj_sample <- no_tj_pitchers |> sample_n(55)

```

Above are my the two random samples I will be using in one of my t-tests to test if there is a significant difference in pitch velocity between pitchers who have had TJ and pitchers who have not. I chose 8 at random for my set.seed and manually add the velocities for all pitchers who showed up as N/A in each sample. I realize now that pitchers were coming up as N/A because I forgot to turn off the setting to only show qualified pitchers when I was taking all of the pitch speed data from online. When I run the same t-test but for the entire populations, I am going to add a line of code that ignores any pitcher without velocity data so that way I don't have to completely re-import all of my data.

### Task 8: Combine random samples
```{r}
sampled_data <- bind_rows(tj_sample, no_tj_sample)
```

I combined my two sets of sampled data so that way I can run the t-test and use the 0 or 1 in the had_tj column to make sure the test is run properly.

### Task 9: T-test with random samples
I am running a t-test with random samples to give me an analysis of the data that is not solely based on the full populations.
```{r}
t_test_sampled <- t.test(ff_avg_speed ~ had_TJ, data = sampled_data)

print(t_test_sampled)
```
This t-test produced a p-value of 0.9453, well above the alpha level of 0.05 where I could confidently reject the null hypothesis. As a result, based on this random sample, I can say that there is not a significant difference in the average fastball velocity of pitchers who have had surgery and pitchers who have not. Therefore, throwing harder most likely does not make a pitcher more likely to need surgery.

### Task 10: T-test with full populations
I am running the same exact two sample t-test as above, just with the full populations for both datasets.
```{r}
t_test_full <- t.test(ff_avg_speed ~ had_TJ, data = velo_for_all, na.rm = TRUE)

print(t_test_full)
```
This t-test produced a p-value of 0.006, well below the alpha level of 0.05, meaning I can confidently reject the null hypothesis. As a result, based on this test, I can say that there is a significant difference in the average fastball velocity of pitchers who have had surgery and pitchers who have not. Therefore, throwing harder most likely does make a pitcher more likely to need surgery.

### Task 11: Logistical regression with full populations
I am running a logisitical regression with the full populations for both datasets.
```{r}
model <- glm(had_TJ ~ ff_avg_speed, family = binomial, data = velo_for_all)

summary (model)
```
This logistical regression produced a p-value of 0.0090, statistically significant at the 99% confidence level, not just 95%. The ff_avg_speed coefficient means that for every 1 mph faster a pitcher throws, the log-odds of surgery increase by 0.123 or approximately 13.1%. 

### Task 12: Logistical regression with random samples
I am running the same exact logisitical regression as above, just with random samples for both datasets.
```{r}
sampled_model <- glm(had_TJ ~ ff_avg_speed, family = binomial, data = sampled_data)

summary(sampled_model)
```
This logistical regression produced a p-value of 0.943, well above the alpha level of 0.05 where I could confidently reject the null hypothesis. Additionally, the ff_avg_speed coefficient means that for every 1 mph faster a pitcher throws, there is actually a slight decrease in a pitchers odds of needing surgery. As a result, based on this random sample, I can say that there is not a significant difference in the average fastball velocity of pitchers who have had surgery and pitchers who have not. Therefore, throwing harder most likely does not make a pitcher more likely to need surgery.

### Regression Curve
I am creating a regression curve that help me visualize the predicted probability of having surgery based on how fast a pitcher throws. I did look up the proper way to format the code to create this graph.
```{r}
velo_for_all$predicted_prob <- predict(model, newdata = velo_for_all, type = "response")
ggplot(velo_for_all, aes(x = ff_avg_speed, y = predicted_prob)) +
  geom_line(color = "red") +
  labs(x = "Fastball Velocity (mph)", y = "Predicted Probability of Surgery") +
  theme_minimal()
```
This regressive curve shows the results of my logistical regression analysis. As pitchers continue to throw faster, their chances of needing Tommy John surgery exponentially increase.

### Multiple Regression
```{r}
other_stats <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQMyEK1XqNnixMrwEwIOKtOuSLeWxebbkxTPGFK3z3i-6CdGTuqTaWMTejJZZEPTKiuXE_jKAgj_yQW/pub?gid=0&single=true&output=csv")
```

```{r}
velo_with_stats <- velo_for_all |>
  left_join(other_stats, by = "Name")
```

```{r}
multi_model <- glm(
  had_TJ ~ IP + FB_pct + CB_pct + SL_pct,
  data = velo_with_stats,
  family = binomial)

summary(multi_model)
```

### Story Pitch
Headline: Data shows MLB pitchers who throw harder are more likely to need Tommy John surgery

My story will explore the link between the average fastball velocity of MLB pitchers and their risk of needing Tommy John surgery (UCL reconstruction). I have analyzed data from 2020 to the present and found that pitchers who throw harder are significantly more likely to require surgery. Major League Baseball has become a league dominated by data and advanced analytics, focused on throwing the ball as hard a possible. This data shows cause for concern, as there is a clear and consistent statistical relationship between higher average fastball velocity and needing surgery.

Main findings: I used a logistical regression to analyze this full set data and found that for each 1 mph increase in average fastball velocity, there is an approximate 13% increase in the odds of needing surgery. This was a statistically significant difference at a 99% confidence level. I also used a two sample test that compared average fastball velocity for pitchers who have and have not had surgery. I found a statistically significant difference in the average speed for these two groups, reinforcing my findings from the regression. I also conducted both of these tests with random samples (about 10% of the full population size) and was surprised to find that they had vastly different results. The random samples presented results that would say there was no statistical significance and that velocity had no correlation with surgery. This showed me that it's harder to detect real differences in small samples, especially with binary datasets (either had TJ or did not). 

Potential problems: I was missing velocity data for several pitchers throughout both of my datasets as a result of the importing process I used. I used baseballsavant.com to get the pitch speed for all of the pitchers and forgot to turn off the setting that only showed qualified pitchers. As a result, some pitchers, especially ones who got injured early in a season, did not have their velocities included. I manually entered the data for any pitcher included in my random samples but had to use na.rm to remove them from my full dataset tests because it would take too long to reimport all of my data. This issue could potentially lead to a slight bias toward active and more established pitchers. There is also a possible issue within selection bias. Pitchers who throw harder might be trusted more by their managers and in turn asked to pitch more innings, leading to overuse, a possible cause of injury. One last potential problem I thought of was the time limitation. Obviously, pitchers have been needings TJ since 1974 when the first operation was conducted. Since I restricted my data to post 2020, it limits my ability to track this trend over a long period of time.

Reporting plan: Assuming I had access and permission, I would interview several pitchers who have had TJ, some who throw hard and some who throw softer, and ask them if they think their velocity played a role in their need for surgery. I would obviously talk to people within the sports medicine industry, especially an orthopedic surgeon, and ask about the mechanics of pitching and if they have seen a link between velocity and UCL damage. Maybe most importantly, I would focus a large chunk of time on the outliers. I would examine as many pitchers who throw hard (likely 97+ mph) as possible who have not had TJ yet. I would look to see if there are doing something different (mechanically, recovery, etc) than pitchers who throw the same speed and have needed surgery. 

All of this reporting combines the statistical evidence with personal and expert insights and will give readers a much deeper understanding of the injury concerns that come with pushing for high velocity in baseball.
