---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: <https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv>. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```

### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 1000, property_rate_2019 = property2019/pop2019 * 1000, violent_rate_2020 = violent2019/pop2020 * 1000, property_rate_2020 = property2020/pop2020 * 1000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

In 2019, Baltimore and Ocean City had the two highest violent crimes rates in the state of Maryland. In 2020, both cities saw their rates barely change, remaining first and second in violent crime rate with 18.9 and 13.1 violent crimes per 1000 people respectively. Looking at property crimes, I noticed that Ocean City also faced a lot of property crimes in 2019, with approximately 104 crimes per 1000 people, the most of any location in Maryland. Baltimore was much lower down in this category, only reporting 42.5 property crimes per 1000 people in 2019. In 2020, Ocean City's property crime significantly increased as they remained in first, but Baltimore's significantly decreased. I think an important story here would be to write how much crime occurs in Ocean City.

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate(violent_pct_change = (violent_rate_2020-violent_rate_2019)/violent_rate_2019 * 100, property_pct_change = (property_rate_2020-property_rate_2019)/property_rate_2019 * 100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

The first thing I noticed when looking at the percent changes was that Baltimore saw property crime decrease by over 25% from 2019 to 2020. I  think that patterns are more interesting than the outliers because they are more representative of the data as a whole and show consistency. The first pattern that caught my eye first was the general trend of property crimes fluctuating significantly more than violent crimes. The biggest change involving violent crime was a 29% increase from 2019-20 in Kitzmiller. However, for property crime, the biggest change was a nearly 320% increase. I also noticed that for both violent crime and property crime, Ocean City had a small percent change from 2019-20 compared to other locations.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the *best* story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Crimes rates continued to increase in Ocean City from 2019-20, data shows.

In both 2019 and 2020, Ocean City, MD, ranked second in violent crime rate, only behind Baltimore. In 2019, Ocean City reported a violent crime rate of 13.1 which stayed steady into 2020. With regards to property crime, Ocean City saw a nearly 17% increase during that same time period, from 103.7 crimes per 1,000 people to 121.3, according to the data. Baltimore, which had the highest violent crime rate both years, did not come close to Ocean City's property crime rate. The data shows that residents of Ocean City may need to start taking extra steps to protect themselves and their property from being victimized.



### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: Percentiles

```{r}
md_cities_percentiles <- md_cities_rates |> mutate(percentile = ntile(violent_rate_2020,100))
```
Not including Princess Anne, MD, which is at exactly the 90th percentile, there are 10 cities in Maryland above the 90th percentile in 2020 violent crime rate.

```{r}
md_cities_percentiles |> summarize(violent_IQR = IQR(violent_rate_2020))
```
While certainly being informative and a helpful statistic, the IQR doesn't change my mind about the best story idea. Instead it actually reinforces my belief that the Ocean City data is the most important. If I were to do the outlier test, I would find that violent crime in Ocean City qualified as one. This proves that crimes are significantly higher than most places in Maryland and definitely are something that residents need to pay attention to.

