---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Gibson Hirt"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

Based on this scatterplot, it appears that there is a relatively strong negative correlation between FARMS percentage and math test scores. It appears that as a schools' FARMS percentage increases, its math test scores decrease. Compared to the scatterplot from the class activity, this graph has obvious correlation. The activity plot has points scattered all over the place and it was obvious there was no pattern within the data. This scatterplot has a distinct shape and based on my preliminary observations, will have a negative correlation.


### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

Provide the model's output for the three statistics mentioned in the class activity.

The model presented a minute p-value of < 0.00000000000000022, an adjusted r-squared value of 0.6734 and an y-intercept of 68.04878 with a FARMS coefficient of -0.59511.

Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

Based on what I've learned so far, these coefficients tell me that there is statistical significance between FARMS percentage and math test scores. The extremely low p-value allows me to reject the null hypothesis and the adjusted r-squared value tells me that 67.34% of the change in math test scores can be explained by FARMS percentage. This model is definitely worth exploring further because there is clearly at least a moderate amount of correlation between the two datasets and we will be able to test if schools are meeting expectations.

### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

The two largest difference between the predicted result and the actual math test scores occurred at Fox Chapel Elementary (29.99) and Broad Acres Elementary (28.48). Fox Chapel is in Germantown, MD, 20876, and the most interesting demographic data I found in this zipcode is that less than 25% of the popualtion is White. The zipcode is just 23% White while it is 28% Black, 21% Asian and 22% Hispanic. Broad Acres is in Silver Spring, MD, 20903, and the most interesting demographic data I found in this zipcode is that over half of the people living in this zipcode are Hispanic (53%). In general, many zipcodes in Montgomery County have predominantly white populations, so it was interesting to see which zipcodes do not fall within this trend.

## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

Based on this model's results, assuming that it is the same school board member from the class activity, I would pitch a story that backs up their claim and explains the conclusions the model has allowed me to come to. I can conclude that the board member is correct in saying that elementary schools with fewer students on free and reduced price meals do better on standardized math tests in Montgomery County, MD. I would be able to point out all of the statistics that prove there is statistical significance between FARMS percentage and math test scores and that FARMS percentage can explain over 65% of the change in math test scores.
